Bayesian Sparsification of $\cplx$-valued networks

### 2020-01-30 paper/paper.tex
-0.5 threshold explanation and a newer abstract

\begin{abstract}
With continual miniaturization ever more applications of deep learning can be found in
embedded systems. Common embedded applications include on-device image recognition and
signal processing (audio and radio). We investigate complex-domain deep neural networks
and complex-valued Bayesian compression thereof. To this end we conduct a large scale
numerical study of complex-valued networks to assess quality - compression rate trade-off.
The results of this study have direct implications for real-time signal processing
throughput of embedded systems.
\end{abstract}

@@

In this section we compare the complex variational dropout techniques discussed above
on the tasks and datasets, that were studied in prior research on complex-valued networks.

% general training strategy
The goal of the following experiments to study the sparsity-performance trade-off of the
proposed Bayesian sparsification methods for $\cplx$-valued networks. To this end we employ
the following training plan in every experiment. The model is trained in three successive
stages: ``dense'' $\to$ ``sparsify'' $\to$ ``fine-tune''.
%
The last two stages are used to compare the .

At the first, ``dense'', stage every network retains its original non-bayesian layers
and is fit ``as-is'' using only the likelihood from \eqref{eq:elbo} without variational
dropout.

During the ``sparsify'' stage we make the relevant layers bayesian and apply $\cplx$-variational
dropout sec.~\ref{ssub:variational_dropout} (or sec.~\ref{ssub:empirical_bayes}) or its $\real$
counterpart. To be able to study the attainable sparsity levels we inject a coefficient $
  C \in (0, 1]
$ at the KL divergence term in \eqref{eq:elbo}:
\begin{equation}  \label{eq:elbo_with_coef}
    - \frac{C}N KL(q_{\theta} \| \pi_{\lambda})
    + \frac1{M} \sum_{i=1}^M
        \log p_{\phi}(z_i \mid g(\varepsilon_i; \theta))
    \,. \tag{\ref{eq:elbo}'}
\end{equation}
Higher $C$ yields increasingly more sparse layers upon termination of this stage.

Before transitioning to the ``fine-tune'' stage we compute the mask of non-zero weights
in each layer based on the relevance scores $\alpha$ (sec.~\ref{sub:dropout}). Since
the variational approximation factorizes into univariate distributions, whether a $\cplx$
or $\real$ parameter is zero or not is determined from its relevance score $
  \alpha = \tfrac{\sigma^2}{\lvert\mu\rvert^2}
$: non-zero (relevant) if $\log \alpha \leq \tau$, zero (irrelevant) otherwise. Figure~%
\ref{fig:typical_log_alpha_distrib} depicts the distribution of $\log \alpha$ values
from \textbf{some run}, which exhibits the typical behaviour we noticed in all experiments
conducted: there always is a single mode within the moderately negative range, and one or
two modes to the right of $0$. The relative mass of each mode changes with $C$ and the
higher the coefficient is, the more mass is shifted to the positive half. However, in every
exepriment the trough between the peaks of the negative mode and the first positive one
is around zero.

Thus, in every experiment we set $\tau = -\tfrac12$, which is admittedly very low, especially
considering that the recommended setting in \citep{molchanov_variational_2017} is $3$. This
yields high levels of sparsity,
The rationale behind our choice was twofold: we use an extra fine tuning
stage, and keep the weights that are away from zero with high probability.

These masks are then used during ``fine-tune'' to train the remaining non-zero parameters
similar to the ``dense'' stage, i.e. with layers non-bayesian layers, but with fixed
sparsity masks.

Unless specified otherwise we use ADAM optimizer with the base learning rate of $10^{-3}$
and the annealing schedule from .

>>>>

\begin{abstract}
With continual miniaturization ever more applications of deep learning can be found in
embedded systems, where it is common to encounter data with natural complex domain
representation. To this end we extend Variational Dropout of \citet{kingma_variational_2015}
and \citet{kharitonov_variational_2018} to complex-valued neural networks building upon
the work of \citet{trabelsi_deep_2017}. We verify the proposed Bayesian technique and
assess the performance-compression trade-off by conducting a large-scale numerical study
of $\cplx$-valued networks on two tasks: image recognition on MNIST-like and CIFAR10
datasets and music transcription on MusicNet \citep{thickstun_learning_2017}. We reliably
replicate the state-of-the-art result for $\cplx$-valued networks on MusicNet with $50-100\times$
compression rate.
\end{abstract}

@@

In this section we compare the complex variational dropout techniques discussed above
on the tasks and datasets, that were studied in prior research on complex-valued networks.
All experiments share the following settings unless specified otherwise:
\begin{itemize}
  \item ADAM optimizer with the base learning rate of $10^{-3}$
  \item learning rate the annealing schedule from \citep{trabelsi_deep_2017}
  \item global gradient clipping to within an $\ell_2$-norm ball of radius $0.05$
\end{itemize}

\subsection{Fitting and compression} % (fold)
\label{sub:fitting_and_compression}

The goal of the following experiments to study the sparsity-performance trade-off of the
proposed Bayesian sparsification methods for $\cplx$-valued networks. To this end we employ
the following training plan in every experiment. The model is trained in three successive
stages: ``dense'' $\to$ ``sparsify'' $\to$ ``fine-tune''. Parameter values and the internal
state of the optimizer for similar parameters are kept between stages.

At the \textit{``dense''} stage every network is fit ``as-is'', retaining its original
non-bayesian architecture and using only the likelihood from \eqref{eq:elbo} without
variational dropout.
%
During the \textit{``sparsify''} stage we make every layer bayesian and apply $\cplx$-variational
dropout sec.~\ref{ssub:variational_dropout} (or sec.~\ref{ssub:empirical_bayes}) or its $\real$
counterpart. To be able to study the attainable sparsity levels we inject a coefficient $
  C \in (0, 1]
$ at the KL divergence term in \eqref{eq:elbo}:
\begin{equation}  \label{eq:elbo_with_coef}
    - \frac{C}N KL(q_{\theta} \| \pi_{\lambda})
    + \frac1{M} \sum_{i=1}^M
        \log p_{\phi}(z_i \mid g(\varepsilon_i; \theta))
    \,. \tag{\ref{eq:elbo}'}
\end{equation}
Higher $C$ yields increasingly more sparse layers upon termination of this stage.
%
At the \textit{``fine-tune''} stage we deduce sparsity masks from the parameters of the fit
variational approximation and train the remaining non-zero parameters similarly to the
``dense'' stage, i.e. with non-bayesian layers, but with fixed sparsity masks.

Between \textit{``sparsify''} and \textit{``fine-tune''} stages we compute masks of non-zero
weights in each layer based on the relevance scores $\alpha$ (sec.~\ref{sub:dropout}). Since
the variational approximation factorizes into univariate distributions, whether a $\cplx$
or $\real$ parameter is zero or not is determined from individual relevance scores $
  \alpha = \tfrac{\sigma^2}{\lvert\mu\rvert^2}
$: non-zero if $\log \alpha \leq \tau$, zero otherwise. Figure~\ref{fig:log-relevance}
(top) displays general shape of the distribution of $\log \alpha$ values from the experiment
in sec.~\ref{sub:mnist_like_datasets}, which depicts the typical behaviour noticed in all
experiments conducted on every considered dataset: there always is a single mode within the
moderately negative range, and one or two modes to the right of $0$. The relative mass of
each mode changes with $C$ and the higher the coefficient is, the more mass is shifted to
the positive half. In every experiment the trough between the peaks of the negative mode
and the first positive one is around zero.

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.9\textwidth}
    \centering
    \includegraphics[width=1.\linewidth,draft]{../notebooks/assets/grad_log.pdf}
    % \caption{} \label{fig:log-relevance-kde}
  \end{subfigure} \\
  % \hfill
  \begin{subfigure}[b]{0.9\textwidth}
    \centering
    \includegraphics[width=1.\linewidth,draft]{../notebooks/assets/grad_log.pdf}
    % \caption{} \label{fig:log-relevance-proba}
  \end{subfigure}
  \caption{%
    \textit{top}: Estimate of $\log\alpha$ density for a convolutional and dense layer
    from the convolutional model of sec.~\ref{sub:mnist_like_datasets}.
    \textit{bottom}: The probability of 
  }
  \label{fig:log-relevance}
\end{figure}

We decide if a parameter is \textit{``non-zero''} if it is within $\delta$ relative
tolerance of its mean with high probability under the approximate posterior. Since
\eqref{eq:c-gauss-vi} is Gaussian, $
  \tfrac{\lvert w - \mu \rvert^2}
        {\alpha \lvert \mu \rvert^2}
$ is $\chi^2_k$ distributed with shape $k=1$ ($\real$) or $2$ ($\cplx$), whence
\begin{equation} \label{eq:prob_relevant}
  \Pr\bigl(
    \lvert w - \mu \rvert \leq \delta \lvert \mu \rvert
  \bigr)
    = \Pr\biggl(
      \frac{\lvert w - \mu \rvert^2}{\alpha \lvert \mu \rvert^2}
        \leq \frac{\delta^2}{\alpha}
    \biggr)
    = \chi^2_k\biggl(\frac{\delta^2}{\alpha}\biggr)
  \,.
\end{equation}
Figure~\ref{fig:log-relevance} (bottom) shows the probability of a $\real$ (or $\cplx$) 
parameter being non-zero for various $\delta\in(0, \tfrac12)$ as a function of $\log \alpha$.
For a reasonably generous $\delta = 25\%$ tolerance all values of $\log \alpha$ below
$-2.5$ yield a reasonably high chance of a non-zero parameter. Given that $q_\theta$ is
merely an approximation we opt for a slightly higher $\log \alpha$ threshold, and thus
set $\tau = -\tfrac12$ in every experiment, despite it being lower than the recommended
$\tau = 3$ in \citep{molchanov_variational_2017}, but yields much higher levels of sparsity.
The rationale behind our choice was twofold: we use an extra fine tuning stage, and keep the
parameters that are away from zero with high probability.

% subsection fitting_and_compression (end)
